### 2025-05-21 basic-statistics.qmd
20250520 - sports analytics

#### Mediation and Moderation

@@

@@

[Descriptive statistics](#sec-descriptiveStatistics) are used to describe the data, including the center, spread, or shape of data.
[Inferential statistics](#sec-inferentialStatistics) are used to draw inferences regarding differences between groups or associations between variables.
Null hypothesis signficance testing is a framework for [inferential statistics](#sec-inferentialStatistics), in which there is a [null hypothesis](#sec-nullHypothesis) and [alternative hypothesis](##sec-alternativeHypothesis).
The null hypothesis is that there is no difference between groups or that there is no association between variables.
[Statistical significance](#sec-statisticalSignificance) is evaluated with a $p$-value, which represents the probability of obtaining a result at least as extreme as the result observed if the null hypothesis is true.
Effects with *p*-values less than .05 are considered statistically significant.
However, it is also important to consider [practical significance](#sec-practicalSignificance) and effect sizes.
When designing a study, it is important to consider [statistical power](#sec-statisticalPower) and the [sample size](#sec-sampleVsPopulation) needed to detect the hypothesized effect size.
You can determine a study's [power](#sec-statisticalPower) based on the [effect size](#sec-practicalSignificance), [sample size](#sec-sampleVsPopulation), and [alpha level](#sec-statisticalSignificance).

>>>>

#### Mediation and Moderation {#sec-statisticalPowerMediationModeration}

@@

## Ethics of Reporting Statistics {#sec-ethicsReportingStatistics}

It is important to report statistics ethically and responsibly.
People can use statistics to lie [@Huff2023].
It is important to present statistics fairly and accurately even if they are not convenient to your preferred narrative.
In addition, it is important to consider [practical significance](#sec-practicalSignificance) in addition to [statistical significance](#sec-statisticalSignificance).

@@

[Descriptive statistics](#sec-descriptiveStatistics) are used to describe the data, including the center, spread, or shape of data.
[Inferential statistics](#sec-inferentialStatistics) are used to draw inferences regarding differences between groups or associations between variables.
Null hypothesis signficance testing is a framework for [inferential statistics](#sec-inferentialStatistics), in which there is a [null hypothesis](#sec-nullHypothesis) and [alternative hypothesis](#sec-alternativeHypothesis).
The null hypothesis is that there is no difference between groups or that there is no association between variables.
[Statistical significance](#sec-statisticalSignificance) is evaluated with a $p$-value, which represents the probability of obtaining a result at least as extreme as the result observed if the null hypothesis is true.
Effects with *p*-values less than .05 are considered statistically significant.
However, it is also important to consider [practical significance](#sec-practicalSignificance) and effect sizes.
When designing a study, it is important to consider [statistical power](#sec-statisticalPower) and the [sample size](#sec-sampleVsPopulation) needed to detect the hypothesized effect size.
You can determine a study's [power](#sec-statisticalPower) based on the [effect size](#sec-practicalSignificance), [sample size](#sec-sampleVsPopulation), and [alpha level](#sec-statisticalSignificance).

### 2025-05-21 cognitive-bias.qmd
20250520 - sports analytics

Focus on the decision-making process when evaluating the quality of decisions.
Yes, also evaluate the outcomes of decisions, but do not give yourself too much credit for lucky decisions or too much blame for unlucky decisions (like injuries).
That said, one thing you can count on is that players get injured.
So, it is to your benefit to draft and compose your team to have some bench depth so you can handle injuries when they inevitably occur.

>>>>

Focus on the decision-making process when evaluating the quality of decisions, rather than the outcome [@Tanney2021].
Yes, also evaluate the outcomes of decisions, but do not give yourself too much credit for lucky decisions or too much blame for unlucky decisions (like injuries).
Luck is going to happen.
If you are making optimal decisions that maximize the expected value of returns (e.g., projected fantasy points), you are following a strong process.
That said, one thing you can count on is that players get injured.
So, it is to your benefit to draft and compose your team to have some bench depth so you can handle injuries when they inevitably occur.

### 2025-05-21 data-visualization.qmd
20250520 - sports analytics

>>>>

Graphics serve an important way of efficiently communicating findings to diverse audiences.
Effective graphics can help people with less technical knowledge better understand your findings.
As noted by @Tanney2021, it is important to communicate your work clearly to stakeholders (e.g., coaches, companies, or prospective clients); if you cannot explain your work clearly, you cannot expect a stakeholder to "buy in" to what you are selling.

### 2025-05-21 decision-making.qmd
20250520 - sports analytics

>>>>

Sports analyst @Tanney2021 encourages people to consider framing your decisions as bets, to use a probabilistic mindset (rather than a determininistic mindset).
Even if something is perceived as likely to occur with 100% certainty, it is not guaranteed that such an event will occur.
Other things can and do happen, and [overconfidence](#sec-cognitiveBiasesOverconfidence) can get in the way of effective decision-making.

A key goal is to identify decisions that provide a net advantage in expected value.
Although there is considerable variability and the outcome is not guaranteed, small advantages can yield considerable gains in the long run @Tanney2021; however, one must be able to survive the costs of losses to obtain enough long-term trials.
So, do not bet more than you are willing to lose.

### 2025-05-21 sports-cognitive-psychology.qmd
20250520 - citation
20250520 - sports analytics

Although baseball was one of the first major sports to embrace analytics, other sports have been transformed by analytics, as well.
For instance, in basketball, there has been a greater focus on three points and buckets close to the rim (e.g., dunks and layups), with way fewer midrange shots [@Partnow2021].
Moreover, star players rest more games.
Here is an article on how Shane Battier, who was not the quickest, fastest, or most athletic player, used analytics successfully to guard players who were quicker or more athletic than him: <https://www.nytimes.com/2009/02/15/magazine/15Battier-t.html> [@Lewis2009b; archived at <https://web.archive.org/web/20250219225627/https://www.nytimes.com/2009/02/15/magazine/15Battier-t.html>].

Football has also seen greater use of analytics, although its uptake has been somewhat slower than in many other sports.
When asked about how he makes the decision about whether to go for a two-point conversion after a touchdown, Steelers Head Coach Mike Tomlin stated:

@@

>>>>

Although baseball was one of the first major sports to embrace analytics, other sports have been transformed by analytics, as well.
For instance, in basketball, there has been a greater focus on three points and buckets close to the rim (e.g., dunks and layups), with way fewer midrange shots [@Partnow2021].
Moreover, star players rest more games ("load management").
Here is an article on how Shane Battier, who was not the quickest, fastest, or most athletic player, used analytics successfully to guard players who were quicker or more athletic than him: <https://www.nytimes.com/2009/02/15/magazine/15Battier-t.html> [@Lewis2009b; archived at <https://web.archive.org/web/20250219225627/https://www.nytimes.com/2009/02/15/magazine/15Battier-t.html>].

Football has also seen greater use of analytics, although its uptake has been somewhat slower than in many other sports [@Carroll2023].
When asked about how he makes the decision about whether to go for a two-point conversion after a touchdown, Steelers Head Coach Mike Tomlin stated:

@@

## Challenging a Call {#challengingCall}

A coach can challenge a call (and have it reviewed) by throwing a red flag onto the field.
If, after the review, the play is reversed, the team does not lose a timeout; however, if the play stands as originally called, the team loses a timeout.
One task is for coaches to determine when it is optimal to challenge a call.
NFL.com provides an opportunity for you to determine whether you should challenge a call: <https://operations.nfl.com/learn-the-game/making-the-call/you-make-the-call/>.

@Tanney2021 provides a framework for determining whether a coach should challenge a call, based on the expected value of the change in winning probability of a reversal versus a non-reversal:

$$
\begin{aligned}
  \text{expected value} &= \text{Sum}(\text{Probabilities} \times \text{Payouts}) \\
  \text{expected value} &= (\text{Reversal \%})(\text{Positive change in Winning Probability of a Reversal}) + (1 - \text{Reversal \%})(\text{Negative change in Winning Probability of Losing a Timeout})
\end{aligned}
$$ {#eq-challengingCall}
